Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-09-08 Chapter_1/04_Manuscript/Manuscript.tex
Various updates

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,\S} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \and
  Gabriele Cozzi\textsuperscript{1,2} \and
  Dominik M. Behr\textsuperscript{1,2}
}

@@

%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
% Importance of Dispersal & Connectivity
Dispersal of individuals is a crucial pre-requisite for long-term species
persistence. However, the ability to disperse is contingent a sufficient degree
of landscape connectivity, which is why the identification and preservation of
movement corridors that promote connectivity has become a task of exceptional
importance.

% Quantifying connectivity
Over the past two decades, ecologists have primarily relied on least-cost
analysis and circuit theory to investigate connectivity. Unfortunately, both
methods make several restrictive assumptions that limit their usefulness in
reality. Least-cost analysis, for instance, assumes that animals move towards a
pre-conceived endpoint and choose a cost-minimizing route accordingly. Circuit
theory, in contrast, presumes a random walk without any directional persistence.
To address these shortcomings, simulations from individual-based movement models
have been proposed and applied, yet a unified framework to simulate dispersal
and to derive connectivity maps based on simulated dispersal events is lacking.

% What we do
% and synthesize a simple
% three-step workflow that uses
Here, we combine several already-existing methods to simulate dispersal and
investigate landscape connectivity in three distinct steps. In the first step,
integrated step selection functions are used to parametrize a mechanistic
movement model. In the second step, the parametrized model is applied as an
individual-based movement model to simulate dispersal trajectories. In the third
step, simulated trajectories are consolidated and synthesized into three
complementary connectivity maps, each focusing on a different aspect of
landscape connectivity. The set of maps includes a heatmap, revealing frequently
traversed areas, a betweenness-map, delineating critical dispersal corridors,
and a map of inter-patch connectivity, depicting the presence or absence of
functional links between distinct habitat patches.

% Case study
As a case study, we apply the proposed workflow to assess landscape connectivity
for dispersing African wild dogs (\textit{Lycaon pictus}) in the Kavango-Zambezi
Transfrontier Conservation Area, the world's largest transboundary conservation
area. With this, we demonstrate that the outlined workflow offers a simple, yet
powerful alternative to traditional connectivity modeling techniques. Although
such individual-based simulations necessitate informed decisions about the
number of simulated individuals and the duration of simulated dispersal events,
they enable to relax several of the unrealistic assumptions inherent to
least-cost analysis or circuit theory.

@@

% Importance of Dispersal & Connectivity
\subsection{Dispersal \& Connectivity Models (90\%)}
Successful dispersal of individuals requires a sufficient degree of landscape
connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the identification
and protection of major dispersal corridors has become a fundamental task in
conservation science \citep{Nathan.2008, Doerr.2011, Rudnick.2012}. To pinpoint
relevant dispersal corridors, a comprehensive understanding of the factors that
limit dispersal of the focal species and an adequate connectivity model are
required \citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the two
most prominent connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}), both graph-based approaches that estimate conductance of the
landscape based on predicted habitat permeability \citep{Zeller.2012}.

% Issues with Both Methods
\subsection{Issues with Traditional Methods (90\%)}
Despite their intuitive nature and ease of use, both methods make several
assumptions that are hardly ever met in reality \citep{Diniz.2020}. With LCPA,
for instance, a least costly route always exists, even if associated movement
costs are unreasonably high and will never be incurred by a dispersing
individual. The method also presumes that animals move towards a pre-conceived
endpoint and choose a cost-minimizing route accordingly. While these assumptions
may by fulfilled by migrating animals, dispersers usually move into unfamiliar
territory and are therefore less likely to be aware of associated movement costs
\citep{Koen.2014, Abrahms.2017, Cozzi.2020}. CT relaxes these assumptions, yet
posits that animals move according to a random walk, entailing that directional
biases cannot be rendered \citep{Diniz.2020}. In reality, however, directional
persistence is regularly observed \citep{Bovet.1991, Schultz.2001}, especially
in dispersing individuals \citep{Cozzi.2020, Hofmann.2021}. Some of these
deficiencies can be addressed using alternative graph-based approaches, such as
least-cost \textit{corridors} \citep{Pinto.2009}, \textit{thresholded}
least-cost paths \citep{Landguth.2012}, or \textit{randomized} least-cost paths
\citep{Panzacchi.2016}, yet a certain degree of arbitrariness remains. Finally,
none of the methods is capable of rendering the temporal dimension of dispersal
\citep{Diniz.2020}, meaning that statements about the expected duration required
to traverse a certain corridor are impossible.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To overcome the unrealistic assumptions inherent to traditional connectivity
modeling techniques, simulations from individual-based movement models (IBMMs)
have been proposed and applied \citep{Diniz.2020}. In these models, dispersal
trajectories are simulated based on movement rules that determine how
individuals move across or interact with the prevailing landscape
\citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004, Revilla.2004,
Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016, Hauenstein.2019,
Zeller.2020, Vasudev.2021}. Instead of enforcing connections towards a known
endpoint, IBMMs allow individuals to choose their route as they move and enable
to render potential interactions between movement behavior and environmental
conditions. IBMMs furthermore enable to model directional persistence and to
study the temporal dimension of dispersal \citep{Diniz.2020}. Despite these
advantages, however, a unified framework to simulate dispersal and to derive
connectivity maps based on simulated dispersal events is lacking.

% The Power of Step Selection Functions
\subsection{Step Selection Functions (90\%)}
A promising framework that has been developed more recently...
Over the past decade, the increased availability of large amounts of data and an
increased computational power have led to the development of numerous techniques
to study animal movement and resource selection during dispersal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Zeller.2012}. Especially
\textit{resource selection functions} \citep{Boyce.2002}, \textit{step selection
functions} \citep{Fortin.2005}, and \textit{path selection functions}
\citep{Cushman.2010} have pushed the field of movement ecology forward. These
methods allow estimating habitat preferences of the focal species by comparing
covariates at locations visited by the animal to the same covariates at
locations available to, but not visited by the animal \citep{Boyce.2002,
Fortin.2005, Cushman.2010, Thurfjell.2014}. While regular SSFs were intended to
learn about relative habitat preferences of the focal species
\citep{Fortin.2005, Thurfjell.2014, Avgar.2017}, the method has recently been
generalized to \textit{integrated} SSFs and now enables to jointly study habitat
and movement preferences, as well as potential interactions between them
\citep{Avgar.2016, Signer.2017, Fieberg.2020}. ISSFs therefore provide a
relatively simple means to model complex movement behavior, where movement is
viewed as the result of two intertwined behavioral kernels (e.g.
\citealp{Prokopenko.2017, Munden.2020}). Importantly, a parametrized ISSF model
represents a fully mechanistic movement model based on which individual movement
trajectories can be simulated \citep{Avgar.2016, Signer.2017}. In fact,
\cite{Signer.2017} used ISSF to simulate steady state utilization distributions
of resident animals that were moving around a point of attraction. However, the
degree to which such simulations are helpful in detecting movement corridors and
modeling landscape connectivity remains to be investigated.

% What we do
\subsection{Proposed Framework (90\%)}
Here, we therefore propose to combine several already-existing methods into a
simple three-step workflow to simulate dispersal and investigate landscape
connectivity. In the first step, integrated step selection functions are used to
parametrize a mechanistic movement model rendering dispersal behavior. In the
second step, the parametrized model is applied as an individual-based movement
model to simulate dispersal trajectories. Finally, in the third step, simulated
trajectories are consolidated into three complementary connectivity maps, each
focusing on a different aspect of landscape connectivity. The set of maps
includes a heatmap, revealing frequently traversed areas, a betweenness-map,
delineating critical dispersal corridors, and a map of inter-patch connectivity,
depicting the presence or absence of functional links between distinct habitat
patches.

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
We exemplify application of the proposed workflow by investigating landscape
connectivity for the endangered African wild dog (\textit{Lycon pictus}). While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range, mainly due to persecution by humans,
habitat fragmentation and destruction, and deadly diseases
\citep{Woodroffe.2012}. Within the few remaining subpopulations, wild dogs form
cohesive packs comprising 8 to 12 adults and their offspring \cite{McNutt.1995}.
After reaching sexual maturity, male and female offspring form same-sex
coalitions and disperse from their natal pack in search for potential mating
partners and a suitable territory to settle \citep{McNutt.1996, Behr.2020}. New
packs are formed when dispersing coalitions join unrelated opposite-sex
dispersing coalitions \citep{McNutt.1996}. During dispersal, wild dogs can cover
several hundred kilometers across a variety of landscapes
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the
main strongholds for this species lies near the Moremi Game Reserve in northern
Botswana, which is part of the world's largest transboundary protected area,
namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA). This
area was originally devoted towards facilitating the migration of elephants, yet
it can be expected to yield benefits to a great multitude of other species
\citep{Elliot.2014, Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
We previously assessed landscape connectivity for dispersing African wild dogs
within the KAZA-TFCA using least-cost methods \citep{Hofmann.2021}. More
specifically, we fitted a habitat selection model and predicted a permeability
surface based on which we computed least-cost paths and corridors between
protected areas. We now expand on this knowledge and use ISSFs to develop a more
detailed, mechanistic movement model of dispersing wild dogs
(\Cref{GraphicalAbstract}). We then employ the model to simulate 80'000
dispersers moving 2'000 steps across the KAZA-TFCA. Finally, we generate a
heatmap, betweenness map and a map of interpatch connectivity. We thereby
exemplify how ISSFs can be utilized for dispersal simulations and we discuss
several benefits of this approach over traditional connectivity modeling
techniques such as least-cost analysis and circuit theory. Most importantly,
simulations based on ISSFs provide a more generic view on how connectivity
emerges and to which degree connectivity depends on the dispersal duration. In
addition, by generating proper dispersal trajectories, network theory can be
applied to calculate network metrics that are pertinent to connectivity
analysis.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_GraphicalAbstract.pdf}
    \caption{Flowchart of the simulation-based connectivity analysis as proposed
    in this article. First, GPS data and habitat covariates have to be
    collected. The combined data is then analyzed in an integrated step
    selection model, which enables the parametrization of the focal species'
    habitat and movement kernels. The parametrized model is then treated as an
    individual-based movement model and used to simulate dispersal trajectories.
    Ultimately, simulated trajectories serve to produce a set of maps that are
    pertinent to landscape connectivity. This includes a heatmap, indicating the
    relative traversal frequency across the study area, a betweenness map,
    highlighting movement corridors and bottlenecks, and, finally, an
    inter-patch connectivity map, where the frequency of connections and their
    average duration can be depicted. \textcolor{red}{Photos: Whom to cite?
    Vectronics or Photographers?}}
    \label{GraphicalAbstract}
  \end{center}
\end{figure}

\section{Methods}
\subsection{Study Area (90\%)}
The study area was defined by a bounding box centered at -17\degree 13'9''S,
23\degree 56'4''E (\Cref{StudyArea}a) stretching over 1.3 Mio.
km\textsuperscript{2} and encompassed the entire KAZA-TFCA (\Cref{StudyArea}b).
The KAZA-TFCA represents the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia. It covers a
total of 520'000 km\textsuperscript{2} and hosts diverse landscapes, ranging
from savanna to grassland and from dry to moist woodland habitats. In its center
lies the Okavango Delta, a dominant hydrogeographical feature and the world's
largest flood-pulsing inland delta. The wet season within the KAZA-TFCA lasts
from November to March and is out of phase with the flood in the Okavango Delta,
which peaks between July and August \citep{McNutt.1996, Wolski.2017}. Although
large portions within the KAZA-TFCA are designated national parks or other
protected areas, considerable human influence remains due to roads, agricultural
sites and settlements and villages that are distributed across the KAZA-TFCA's
landscape.

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
        \includegraphics[width=\textwidth]{99_StudyArea.png}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % % next four lines will help you to locate the point needed by forming a grid.
            % % comment these four lines in the final picture.
            % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
            % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
            % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
            % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
            % % upto here
            \draw[densely dotted, black] (0.169, 0.222) -- (0.364, 0.955);
            \draw[densely dotted, black] (0.169, 0.157) -- (0.364, 0.074);
        \end{scope}
    \end{tikzpicture}
    \caption{Illustration of the study area located in southern Africa. (a) The
    study area was confined by a bounding box spanning the entire KAZA-TFCA and
    encompassed parts of Angola, Namibia, Botswana, Zimbabwe, and Zambia. (b)
    The KAZA-TFCA represents the world's largest terrestrial conservation area
    and covers a total of 520'000 km\textsuperscript{2}. Its purpose is to
    re-establish connectivity between already-existing national parks (dark
    green) and other protected areas (light green).}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data (90\%)}
GPS data of 16 dispersing wild dogs was obtained as described in
\citep{Cozzi.2020, Hofmann.2021}. We collected data on 7 female and 9 male
dispersing coalitions between 2011 and 2019 from a free-ranging population
inhabiting the Moremi National Park in northern Botswana. In some instances,
exact dispersal dates were known from field observations, in other cases we
determined dispersal phases using the net-squared displacement metric. Net
squared displacement measures the squared Euclidean distance of a GPS relocation
to a reference point \citep{Borger.2012}, which in our case was set to the
center of each individual's natal home range. Thus, dispersal was deemed to have
started when an individual left its natal home range and ended once individuals
became sedentary again. Because behavior during dispersal is more pertinent to
landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded all data
that was collected during residency. During dispersal, collars were programmed
to record a GPS fix every four hours and regularly transmitted over the Iridium
satellite system. As previous research found no differences in behaviors of
females and males during dispersal \citep{Woodroffe.2019, Cozzi.2020}, we did
not distinguish between sexes. After collection, we converted GPS coordinates (n
= 4'169) to steps, where each step represented the straight-line distance
traveled by an individual between two consecutive GPS relocations
\citep{Turchin.1998}. To ensure a regular sampling interval, we removed fixes
that were not successfully collected on the 4-hourly schedule (\( \pm \) 15
minutes).

\subsection{Covariates (90\%)}
We represented the physical landscape across the study area using a set of
habitat covariates that included water-cover, distance to water, woodland-cover,
and shrub/grassland-cover. Because water cover greatly changes within and
between years in the Okavango Delta, we applied a remote sensing algorithm and
generated frequently updated water cover layers and corresponding distance to
water layers (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting layers thus temporally aligned with each
dispersal event, ensuring minimal temporal lag between observed movements and
corresponding water layer. We furthermore computed a proxy for human influence,
rendering anthropogenic pressures stemming from human-density, agricultural
sites, and roads. All spatial layers were coarsened or interpolated to a target
resolution of 250 m by 250 m. A detailed description of the preparation of each
habitat covariate is given in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}) (for details see \citep{Avgar.2016, Fieberg.2020}). Because
wild dogs follow a diurnal activity pattern \citep{Castello.2018}, we also coded
a binary variable (\textsf{LowActivity}) indicating whether a step was realized
during periods of low wild dog activity (17:00 to 09:00 local time) or high wild
dog activity (09:00 to 17:00 local time). Handling and manipulation of all data,
as well as all models and simulations were implemented with the statistical
software {\tt R}, version 3.6.6 \citep{R.2019}. Several helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}.

>>>>

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \and
  Gabriele Cozzi\textsuperscript{1,2} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \and
  Dominik M. Behr\textsuperscript{1,2}
}

@@

%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
% Importance of dispersal & connectivity and mow its measured
Dispersal of individuals is a crucial prerequisite for long-term species
persistence but depends on a sufficient degree of landscape connectivity. To
date, connectivity has primarily been quantified using least-cost analysis and
circuit theory, albeit both methods make assumptions that are rarely met in
reality. Least-cost analysis assumes that animals move towards a preconceived
destination and choose a cost-minimizing route accordingly, whereas circuit
theory presumes a complete random walk without directional persistence. While
these shortcomings can be overcome by spatio-temporally explicitly simulating
dispersal trajectories from individual-based movement models, a unified
approach for such simulations is lacking.

% What we do
Here, we present a simple three-step workflow to simulate dispersal and asses
landscape connectivity starting from empirical GPS movement data. We exemplify
the workflow using dispersal data collected on the endangered African wild dog
(\textit{Lycaon pictus}) in the Kavango-Zambezi Transfrontier Conservation Area,
the world's largest transboundary conservation area. In step one, we use
integrated step selection functions to parametrize a mechanistic movement model
describing habitat and movement preferences of dispersers. In step two, we
employ the parametrized model to simulate individual dispersal trajectories
across the landscape. In step three, we interpret simulated trajectories by
converting them into three complementary connectivity maps: a heatmap revealing
frequently traversed areas, a betweenness-map delineating dispersal corridors,
and a map of inter-patch connectivity depicting the presence and intensity of
functional links between habitat patches.

% Some results
We find that simulating dispersal from indvidual-based movement models fitted
via step selection functions are readily interpretable and offer great
flexibility to realistically render dispersal movements. Moreover, by not
enforcing connections towards known endpoints but instead letting dispersers
react to the environment as they go, such simulations overcome shortcomings
inherent least-cost path analysis, while at the same time allowing for
autocorrelation in simulated steps, which can not be incorporated using circuit
theory. Inspite of these advantages, simulating dispersal entails numerous
non-trivial modeling decisions, such as the number of simulated individuals and
the duration of simulated dispersal events.

% Case study
We  show that simulations from step-selection functions offer a simple yet
powerful alternative to traditional connectivity modeling techniques, although
necessitating informed decisions about the number of simulated individuals and
the duration of simulated dispersal events. Ultimately, our workflow permits a
more mechanistic understanding of dispersal behaviour and landscape connectivity
making it useful for a variety of applications in ecological, evolutionary, and
conservation research.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
or recolonize unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However,
the ability to disperse is contingent on a sufficient degree of landscape
connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the identification
and protection of dispersal corridors that promote connectivity has become a
task of fundamental importance \citep{Nathan.2008, Doerr.2011, Rudnick.2012}.
The identification of dispersal corridors not only necessitates a comprehensive
understanding of the factors that limit dispersal, but also an appropriate model
to estimate connectivity \citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To
date, the two most prominent connectivity models in the literature are
least-cost path analysis (LCPA; \citealp{Adriaensen.2003}) and circuit theory
(CT; \citealp{McRae.2006, McRae.2008}), both graph-based approaches that
quantify conductance of the landscape based on habitat permeability
\citep{Zeller.2012}.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Despite their intuitive nature and ease of use, both methods make rigorous
assumptions that are hardly ever met in reality \citep{Diniz.2019}. With LCPA,
for instance, a least-costly route always exists, even if associated movement
costs are unreasonably high and will never be incurred by dispersing individuals
\citep{Sawyer.2011}. The method also presumes that animals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Abrahms.2017}. While these assumptions may be fulfilled by migrating
animals, they are unlikely to hold for dispersers, as dispersers typically move
across unfamiliar territory and are therefore less aware of associated movement
costs \citep{Koen.2014, Cozzi.2020}. CT relaxes these assumptions, but posits
that animals move according to a random walk, entailing that autocorrelation
between subsequent movements cannot be rendered \citep{Diniz.2019}. In reality,
however, autocorrelation in animal movements is regularly observed
\citep{Bovet.1991, Schultz.2001}, especially during dispersal \citep{Cozzi.2020,
Hofmann.2021}. Some of these deficiencies can be alleviated using less stringent
graph-based methods, such as least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, or
\textit{randomized} least-cost paths \citep{Panzacchi.2016}, yet a certain
degree of arbitrariness remains. Besides this, graph-based methods are incapable
of rendering the temporal dimension of dispersal, meaning that statements about
the approximate duration required to move between habitats are impossible
\citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
To overcome the unrealistic assumptions of LCPA and CT, dispersal simulations
from individual-based movement models (IBMMs) have been proposed and applied
\citep{Diniz.2019}. In these models, dispersal trajectories are simulated based
on movement rules that govern how individuals move across and interact with the
prevailing landscape (e.g. \citealp{Gustafson.1996, Gardner.2004, Graf.2007,
KramerSchadt.2004, Revilla.2004, Revilla.2008, Kanagaraj.2013, Clark.2015,
Allen.2016, Hauenstein.2019, Zeller.2020, Vasudev.2021}). Once the trajectories
are simulated, they are further processed and converted into measures of
connectivity \citep{Kanagaraj.2013, Zeller.2020}. Because IBMMs do not enforce
connections towards preconceived endpoints, simulated individuals are allowed to
dynamically adjust their behavior as they disperse. This enables to render
potential interactions between movement behavior and habitat conditions,
shifting the focus from a structural to a more functional view on connectivity
\citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}. Furthermore, IBMMs
render movement sequentially, meaning that the temporal dimension of dispersal
becomes explicit and that autocorrelation between movements can be incorporated
\citep{Diniz.2019}. In spite of these advantages, a unified framework to
simulate dispersal and assess connectivity using IBMMs is lacking.

% Proposed Solution: Three-Step Workflow
\subsection{Proposed Solution: Three-Step Workflow}
In this paper, we draw on recent innovations in the field of movement ecology
and network theory to propose a simple three-step workflow for the simulation of
dispersal and assessment of connectivity (\Cref{GraphicalAbstract}). In the
first step, GPS data of the focal species is combined with relevant habitat
covariates and analyzed in an integrated step selection framework to fit a
movement model. In the second step, the parametrized model is employed as an
IBMM to simulate dispersal trajectories across the study area. Finally, in the
third step, simulated trajectories are converted into a set of three
complementary connectivity maps, each focusing on a different aspect of
connectivity. The set of maps includes a heatmap, revealing frequently traversed
areas, a betweenness-map, delineating dispersal corridors and bottlenecks, and a
map of inter-patch connectivity, depicting the presence or absence, intensity,
and dispersal duration of functional links between habitat patches.

% Some Background on SSFs
\subsection{Some Background on SSFs}
To fit a movement model, we propose to employ the framework of step selection
functions (SSFs; \citealp{Fortin.2005, Thurfjell.2014, Avgar.2016}). Originally,
SSFs were developed to learn about habitat preferences of the focal species
using GPS data \citep{Fortin.2005}. The method works by comparing covariates
along realized steps (the straight-line connecting two consecutive GPS
relocations; \citealp{Turchin.1998}) to the same covariates along randomly
generated alternative steps \citep{Fortin.2005, Thurfjell.2014} using
conditional logistic regression \citep{Fortin.2005, Muff.2020}. Recently,
\cite{Avgar.2016} have introduced \textit{integrated} SSFs (ISSFs), which is a
generalization of regular SSFs and requires that movement metrics are included
in the corresponding regression model. This not only reduces biases in estimated
preferences, but also permits simultaneous inference on a species' habitat
kernel (i.e. habitat preferences), its movement kernel (i.e. movement
preferences/capabilities), and potential interactions between the two kernels
\citep{Avgar.2016, Fieberg.2020}. Importantly, a model that was fitted using
ISSFs resembles a fully mechanistic IBMM, based on which movement can be
simulated \citep{Avgar.2016, Signer.2017}. Previously, ISSFs have been applied
to simulate utilization distributions of resident animals \citep{Avgar.2016,
Signer.2017}, yet the method remains underutilized for the simulation of
dispersal and detection of movement corridors.

% Some Background on Connectivity "Metrics"
\subsection{Some Background on  Connectivity ``Metrics'' }
Irrespective of the chosen modeling framework, simulated dispersal trajectories
need to be processed into measures of connectivity. To date, the most common
practice to translate simulations into connectivity measures is to compute
inter-patch connectivity between distinct habitats \citep{Diniz.2019}.
Inter-patch connectivity is a simple ratio between the number of dispersers
successfully moving between habitat patches and the total number of simulated
individuals from each patch (e.g. \citealp{Kanagaraj.2013}). As an alternative,
some researchers have tallied overlapping trajectories into what might be called
``heatmaps'', i.e. maps that depict the frequency at which each spatial unit in
the study area was traversed by simulated dispersers (e.g. \citealp{Peer.2008,
Hauenstein.2019, Zeller.2020}). More recently, \cite{BastilleRousseau.2018}
proposed a method to convert trajectories into a network, so that metrics from
network theory can be computed. With regards to connectivity, the betweenness
metric seems promising, as it indicates the number of shortest paths going
through each node in the network and therefore highlights areas of exceptional
importance for connecting different regions in the study area. Because each of
the three metrics metrics focuses on a very different aspect of landscape
connectivity, we propose to view them as complementary and to use them in
concert when assessing connectivity.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_GraphicalAbstract2.pdf}
    \caption{Flowchart of the simulation-based connectivity analysis as proposed
    in this article. First, GPS data and habitat covariates have to be
    collected. The combined data is then analyzed in an integrated step
    selection model, which enables the parametrization of the focal species'
    habitat and movement kernels and results in a mechanistic movement model.
    The parametrized model is then treated as an individual-based movement model
    and used to simulate dispersal trajectories. Ultimately, simulated
    trajectories serve to produce a set of maps that are pertinent to landscape
    connectivity. This includes a heatmap, indicating the traversal frequency
    across each spatial unit of the study area, a betweenness map, highlighting
    movement corridors and bottlenecks, and, finally, an inter-patch
    connectivity map, where the frequency of connections and their average
    duration can be depicted.}
    \label{GraphicalAbstract}
  \end{center}
\end{figure}

% Case Study
\subsection{Case Study}
To exemplify the application of the proposed workflow in
(\Cref{GraphicalAbstract}), we investigate landscape connectivity for the
endangered African wild dog (\textit{Lycon pictus}) as a case study. While once
present across entire sub-Saharan Africa, this species has disappeared from a
vast majority of its historic range, mainly due to persecution by humans,
habitat fragmentation and destruction, and deadly diseases
\citep{Woodroffe.2012}. Some of the biggest remaining populations reside near
the Moremi Game Reserve in northern Botswana \citep{McNutt.1996, Cozzi.2013,
Behr.2020}, which is part of the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA), the world's largest transboundary protected area. We previously
assessed landscape connectivity for dispersing African wild dogs within this
ecosystem using least-cost methods \citep{Hofmann.2021}. More specifically, we
used ISSFs to fit a habitat selection model and predicted a permeability surface
based on which we computed least-cost paths and corridors between protected
areas. We now expand on this knowledge and use ISSFs to develop a more detailed,
mechanistic movement model of dispersing wild dogs (\Cref{GraphicalAbstract}).
We then employ the model to simulate 80'000 dispersers moving 2'000 steps across
the landscape of the KAZA-TFCA. Finally, we condense the simulated trajectories
into a heatmap, betweenness-map, and inter-patch connectivity map. We thereby
demonstrate that simulations from ISSFs provide a viable alternative to LCPA or
CT and deliver a more comprehensive understanding of the emergence of
connectivity. Nevertheless, we recognize that simulations from ISSFs entail
several non-trivial modeling-decisions, such as the number of simulated
dispersers and the duration of simulated dispersal events.

\section{Methods}
\subsection{Study Area}
The study area of our wild dog case study was defined by a bounding box centered
at -17\degree 13'9''S, 23\degree 56'4''E (\Cref{StudyArea}a) and stretched over
1.3 Mio. km\textsuperscript{2}, encompassing the entire KAZA-TFCA
(\Cref{StudyArea}b). The KAZA-TFCA comprises parts of Angola, Botswana, Namibia,
Zimbabwe, and Zambia and hosts diverse landscapes, ranging from savannah to
grassland and from dry to moist woodland habitats. In its center lies the
Okavango Delta, a dominant hydrogeographical feature and the world's largest
flood-pulsing inland delta. Although large portions of the KAZA-TFCA are
designated national parks and other protected areas, considerable human
influence remains from roads, agricultural sites and settlements.

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
        \includegraphics[width=\textwidth]{99_StudyArea.png}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % % next four lines will help you to locate the point needed by forming a grid.
            % % comment these four lines in the final picture.
            % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
            % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
            % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
            % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
            % % upto here
            \draw[densely dotted, black] (0.169, 0.222) -- (0.364, 0.955);
            \draw[densely dotted, black] (0.169, 0.157) -- (0.364, 0.074);
        \end{scope}
    \end{tikzpicture}
    \caption{Illustration of the study area located in southern Africa. (a) The
    study area was confined by a bounding box spanning the entire KAZA-TFCA and
    comprised parts of Angola, Namibia, Botswana, Zimbabwe, and Zambia. (b)
    The KAZA-TFCA represents the world's largest terrestrial conservation area
    and covers a total of 520'000 km\textsuperscript{2}. Its main purpose is to
    re-establish connectivity between already-existing national parks (dark
    green) and other protected areas (light green).}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Data}
GPS data of 16 dispersing wild dogs was obtained as described in
\cite{Cozzi.2020} and \cite{Hofmann.2021}. We collected data on 7 female and 9
male dispersers between 2011 and 2019 from a free-ranging population inhabiting
the Moremi National Park in northern Botswana. Because behavior during dispersal
is more pertinent to landscape connectivity than behavior during residence
\citep{Elliot.2014, Abrahms.2017}, we discarded all data that was collected
during residency. For some individuals, exact dispersal dates were known from
field observations, whereas in other cases, we determined dispersal phases using
net-squared displacement. Net-squared displacement is a metric that measures the
squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which in our case was set to the center of each
individual's natal home range. Thus, dispersal was deemed to have started when
an individual left its natal home range and ended once individuals became
sedentary again. During dispersal, GPS collars recorded a fix every 4 hours and
they regularly transmitted data over the Iridium satellite system. After
collection, we converted GPS coordinates of dispersers (n = 4'169) to steps,
where each step represented the straight-line connecting two consecutive GPS
relocations \citep{Turchin.1998}. To ensure a regular sampling interval, we
removed fixes that were not successfully collected on the 4-hourly schedule (\(
\pm \) 15 minutes) and identified bursts of regularly sampled GPS fixes.

\subsection{Covariates}
We represented the physical landscape in our study area using a set of habitat
covariates, including water-cover, distance-to-water, woodland-cover, and
shrub/grassland-cover. Because water-cover greatly changes within and between
years in the Okavango Delta, we applied a remote sensing algorithm and obtained
weekly updated water-cover layers and corresponding distance-to-water layers
from MODIS satellite imagery (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting layers thus temporally aligned with each
dispersal event, ensuring minimal temporal lag between observed movements and
corresponding water layers. We furthermore included a proxy for human influence,
rendering anthropogenic pressures stemming from human-density, agricultural
sites, and roads. All spatial layers were coarsened or interpolated to a target
resolution of 250 m by 250 m. A detailed description of the preparation of each
habitat covariate is provided in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in our ISSF models \citep{Avgar.2016, Fieberg.2020}.
Movement metrics were calculated for each step and included the step length
(\textsf{sl}), its natural logarithm (\textsf{log(sl)}), and the cosine of the
relative turning angle (\textsf{cos(ta)}). Because wild dogs follow a distinct
activity pattern \citep{Castello.2018}, we also coded a binary variable
(\textsf{LowActivity}) indicating whether a step was realized during periods of
low wild dog activity (09:00 to 17:00 local time) or high wild dog activity
(17:00 to 09:00 local time). Handling and manipulation of all data, as well as
all models and simulations were implemented with the statistical software {\tt
R}, version 3.6.6 \citep{R.2020}. Some helper functions were written in {\tt
C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}.
